Ornaments as indicators of social changes and cultural practices in northeast Taiwan before and after European colonial period

### 2019-11-29 analysis/paper/paper.Rmd
added sentences about origin of beads and metal ornaments

---
title: "Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1400-1900 AD) in northeastern Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from a colonial influence. Ornaments give insights into the increasing social inequality stimulated by European colonisation, and show the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonisation; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Consumption patterns of foreign goods can give insights into negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000; @Voss2005]. Northeastern Taiwan is an ideal context to study peripheral colonial influence because although there was a prominent Spanish and Dutch colonial presence in Taiwan, the northeast region was isolated from intensive direct contact by the Xueshan Mountains.

This article describes ornaments excavated from Kiwulan (700-1900 AD), the largest Iron Age settlement on the Yilan plain in northeast Taiwan. The first recorded European presence in Yilan was a Spanish revenge attack on indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rules and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-2, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, I explore archaeological ornaments from Kiwulan spanning the pre-European contact period, the period of Spanish and Dutch presence, and the period of Chinese presence. I address the question of whether indirect colonial influences on the indigenous populations can be detected through the ornament assemblages. 
# Historical background

The island of Taiwan lies at the junction of mainland China, southeast Asia, and northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered historical period since the early 17th century due to the colonial presence of the Spanish and Dutch who played an important role in keeping written records about Taiwan. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged agate beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of agate beads in ritual contexts, with more agate beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that agate beads were already treated as prestige goods in Yilan before the arrival of Europeans.

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232].

Recent ethnographic research with the members of the Kavalan tribe, the contemporary indigenous people of the Yilan plain, describes how agate beads were used in divination practices by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as heirlooms in shaman families. When not in use, beads were often hidden in safe places, such as on the roof beams. Despite most Kavalan people today not knowing the specific origin of their agate beads, the high value of beads is still indicated in oral history. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. However, compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.

@@

The ornaments found in the Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), agate beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table). The chemical composition analysis for glass beads shows high-lead content in the glass, coupled with the winding/folding technique of bead making, suggest a China production tradition[@Cheng2008; @Wang2018]. The style of other ornaments, including glass beads, metal bells, agate beads, are also associate with a Chinese origin [@Chen2007]. There is no clear evidence shows the European origin of beads, however, the large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans via economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007].

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequencies of the major classes of ornaments for different the time periods at Kiwulan. The difference in frequencies between the three time periods reflect significant differences in the use of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$). Most ornament types were present before European contact. Ornament frequencies reached a peak during the European period and then dropped during the Chinese period, especially golden beads. This trend can be also seen on other ornaments including agate beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European period and a further decrease in the Chinese period.

@@

The distribution of frequencies for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. Agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for agate beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the European presence. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in clusters that indicate their original use as strings or necklaces. Copper bells are typically >2 cm long with a wide variety of human faces as a motif, while glass beads are usually <0.5 cm long Indo-Pacific beads, widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

Figure \@ref(fig:plot-all-spatial-pattern) presents the spatial distribution of all ornaments from the research area for each time period. Before the European arrival, a greater amount of ornaments were found at the northern and middle parts of the research area. During the European period, ornaments were more widespread, with some clusters on the northern part. During the Chinese period the distribution is more even again. Figure \@ref(fig:plot-spatial-pattern-each) presents the distribution for the major ornament classes individually, some clusters across the area can be observed during the European period, such as golden beads and agate beads. However, there seems to be no clear consistent pattern across those different ornaments. Each class shows its own pattern where the squares with higher number of ornaments distributed separately and independently. For example, a cluster of golden beads was found at the northern part, while a cluster of agate beads was found in the middle part. In contrast, there are multiple clusters of metal rings that are distributed separately across the research area. Copper bells were usually found individually and appear randomly distributed across the area. In the Chinese period, both the amount and density of different classes of ornaments decreased.

>>>>

---
title: "Ornaments as indicators of social changes and cultural practices in northeast Taiwan before and after European colonial period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1400-1900 AD) in northeast Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from a colonial influence. Ornaments give insights into the increasing social inequality stimulated by European colonisation, and show the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonisation; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Consumption patterns of foreign goods can give insights into negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000; @Voss2005]. Northeast Taiwan is an ideal context to study peripheral colonial influence because although there was a prominent Spanish and Dutch colonial presence in Taiwan, the northeast region was isolated from intensive direct contact by the Xueshan Mountains.

This article describes ornaments excavated from Kiwulan (700-1900 AD), the largest Iron Age settlement on the Yilan plain in northeast Taiwan. The first recorded European presence in Yilan was a Spanish revenge attack on indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rules and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-1662, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, I explore archaeological ornaments from Kiwulan spanning the pre-European contact period, the period of Spanish and Dutch presence, and the period of Chinese presence. I address the question of whether indirect colonial influences on the indigenous populations can be detected through the ornament assemblages. 
# Historical Background of complex exchange network during the late Iron Age and early historical period

The island of Taiwan lies at the junction of mainland China, southeast Asia, and northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered historical period since the early 17th century due to the colonial presence of the Spanish and Dutch who played an important role in keeping written records about Taiwan. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use carnelian as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use carnelian beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that agate beads were already treated as prestige goods in Yilan before the arrival of Europeans.

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers. However, we must recognize some biases in historical records that might lead to misinterpretations of the use of ornaments. Inconsistent descriptions from different written sources and a lack of context in texts are the main problems need to be resolved by studying the ornaments in archaeological contexts.

Currently the ornaments found in northeast Taiwan in early historical period around 400 hundred years ago, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metal working, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is wide varity of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicate multiple origins mostly likely from Southeast Asia [@Chen]. There is no clear evidence shows the European origin of beads, however, the large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Northeast Taiwan from the late Iron age to early historical period was involved in complex exchange network both in a regional and global scale.

@@

The ornaments found in the Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), carnelian beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table).

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequencies of the major classes of ornaments for different the time periods at Kiwulan. The difference in frequencies between the three time periods reflect significant differences in the use of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$). Most ornament types were present before European contact. Ornament frequencies reached a peak during the European period and then dropped during the Chinese period, especially golden beads. This trend can be also seen on other ornaments including carnelian beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European period and a further decrease in the Chinese period.

@@

The distribution of frequencies for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the European presence. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in clusters that indicate their original use as strings or necklaces. Copper bells are typically >2 cm long with a wide variety of human faces as a motif, while glass beads are usually <0.5 cm long Indo-Pacific beads, widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

Figure \@ref(fig:plot-all-spatial-pattern) presents the spatial distribution of all ornaments from the research area for each time period. Before the European arrival, a greater amount of ornaments were found at the northern and middle parts of the research area. During the European period, ornaments were more widespread, with some clusters on the northern part. During the Chinese period the distribution is more even again. Figure \@ref(fig:plot-spatial-pattern-each) presents the distribution for the major ornament classes individually, some clusters across the area can be observed during the European period, such as golden beads and carnelian beads. However, there seems to be no clear consistent pattern across those different ornaments. Each class shows its own pattern where the squares with higher number of ornaments distributed separately and independently. For example, a cluster of golden beads was found at the northern part, while a cluster of carnelian beads was found in the middle part. In contrast, there are multiple clusters of metal rings that are distributed separately across the research area. Copper bells were usually found individually and appear randomly distributed across the area. In the Chinese period, both the amount and density of different classes of ornaments decreased.
